---
title: "Daily Assignments 27 and 28"
subtitle: 'Ecosystem Science and Sustainability 330'
author:
  - name: Samantha Nauman
    email: "snauman4@colostate.edu"
format: html
execute:
  echo: true
---
```{r}
# Libraries
library(osmdata)
library(sf)
library(terra)
library(dplyr)
library(ggplot2)
library(tidyverse)
```
## 1. Define a river object
```{r}
# Create a geocode
fc = AOI::geocode("Fort Collins", bbox = TRUE)

# Get a river object 
foco_rivers <- osmdata::opq(st_bbox(st_transform(fc,4326))) |> 
  add_osm_feature("waterway") |> 
  osmdata_sf()
```

## 2. Filter for Cache la Poudre River and Merge Lines
```{r}
# Filter to Cache la Poudre River
poudre <- foco_rivers$osm_lines %>%
  filter(str_detect(name, "Cache la Poudre"))

# Merge into single line
poudre_river <- st_union(poudre) |> 
  st_as_sf()
```

## 3. Calculate River Length
```{r}
river_length <- st_length(poudre_river)
print(river_length)
```

The river length is 175,417.7 meters.

## 4.  Convert to POINTS
```{r}
poudre_pts <- st_cast(poudre_river, "POINT")
```

## 5. Define DEM Object
```{r}
# Fixes to make the dem code work 
url <- 'https://raw.githubusercontent.com/mikejohnson51/csu-ess-330/a993e03df117a76c609ff4c018055f8c821a6de9/resources/foco-elev.tif'

rast(glue::glue("/vsicurl/{url}"))

Sys.setenv("AWS_NO_SIGN_REQUEST" = "YES")
rast("/vsis3/lynker-spatial/gridded-resources/dem.vrt")

# Load DEM from S3 (adjust path if needed)
dem <- rast('/vsis3/lynker-spatial/gridded-resources/dem.vrt')
```

## Extract River Profile (6,7,8)
```{r}
# Extract elevation from DEM
elevations <- terra::extract(dem, vect(poudre_pts))

# Combine with geometry
river_profile <- bind_cols(poudre_pts, elevations) |>
  mutate(ID = row_number())
```
## Compute Sinuosity (9,10)
```{r}
start_end_dist <- st_distance(poudre_pts[1,], poudre_pts[nrow(poudre_pts),])
sinuosity <- as.numeric(river_length) / as.numeric(start_end_dist)
print(sinuosity)
```
The sinuosity is 2.99028. This indicates that the Cache la Poudre River meanders a lot rather than going straight, meaning it has numerous bends and curves.  

## 11. Compute Slope
```{r}
elev_change <- river_profile$dem[1] - river_profile$dem[nrow(river_profile)]
slope <- as.numeric(elev_change) / as.numeric(river_length)
print(slope)
```
The slope of the river is 0.9781622, meaning the river has some steep sections, but also some gentler stretches. 

## 12. Elevation Profile Plot
```{r}
ggplot(river_profile, aes(x = ID, y = dem)) +
  geom_line(color = "blue") +
  labs(x = "River Point ID", y = "Elevation (cm)", title = "Cache la Poudre River Elevation Profile") +
  theme_minimal()
```

## 13. Map of River Colored by Elevation
```{r}
ggplot(river_profile) +
  geom_sf(aes(color = dem)) +
  scale_color_viridis_c() +
  labs(title = "River Map Colored by Elevation", color = "Elevation (cm)") +
  theme_minimal()
```
```